Variation in gaze understanding across the life span: A process-level perspective

### 2023-09-18 paper/manuscript.Rmd
complete first draft

abstract: |
  Observing eye gaze is fundamental for many social-cognitive abilities, for example, when judging what another agent can or cannot know. While the emergence of gaze following in infancy has been thoroughly studied, we know little about the developmental trajectory throughout childhood and adolescence. The present study examines gaze understanding across the lifespan. We formalize the process of gaze understanding in a computational cognitive model that allows us to conceptualize individual differences in a psychologically meaningful way. The key parameter in our model is an inferential component, which describes how accurately a participant infers a target location based on the agent's gaze. We test a fundamental assumption of our gaze model by studying vector estimation in a non-social setting. Lastly, we explore the relationship between other social-cognitive abilities and gaze understanding. We found a steep learning curve in gaze understanding during the preschool years. In these years, children become more precise in locating the attentional focus of an agent. Precision levels then stay comparably stable, with a minor decay toward older adulthood. Our gaze model estimates reliable individual parameter values and recovers signature patterns in the data. Furthermore, we found that gaze understanding is associated with non-social vector estimation and perspective-taking but not with other Theory of Mind tasks. This work illustrates how the combination of reliable tasks and formal theoretical models allows us to explore the in(ter)dependence of core social-cognitive processes in greater detail.
keywords          : "social-cognitive development, theory of mind, gaze cues, individual differences, cognitive modeling, lifespan"
wordcount         : "xxx max 8000"
floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes 
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
header-includes   :
  - \usepackage{setspace}
  - \captionsetup[figure]{font={stretch=1}}
editor_options    : 
  markdown: 
    wrap: sentence
bibliography      : "../../references.bib"
---

@@

At first glance, the debate about infants’ gaze-following abilities seems to be settled. On further view, however, some methodological choices in the experimental paradigms as well as the scope of infants’ abilities seem to offer room for further exploration. We see the following challenges: (1) Studies focus on infancy, leaving aside further potential developments in childhood and adolescence; (2) Gaze cues are often accompanied by a head movement in the same direction, which makes it hard to isolate the signaling power of the eyes; (3) Often, infants’ looking times are taken as a benchmark for passing a task, while it is uncertain how this would translate into active behavior; (4) Presenting two objects drastically limits the complexity of the visual input, especially when compared to our daily, visually cluttered environment; (5) Behavior is grouped into passing/failing or a comparison against chance, ignoring the accuracy of children; (6) Analyses focus on group-level averages, overlooking variation in the sample.

Adding to the first point, while the emergence of gaze following has been well established, less is known about the developmental trajectory throughout childhood and adolescence. One possibility is that our social-cognitive ability in question is fully developed once emerged in infancy. However, many cognitive abilities develop with age [e.g., working memory, @gathercole2004structure]. Similarly, visual processing appears to improve with age. Therefore, children could potentially improve in gaze following, fine-tuning the performance of the already existing skill. We believe that the development of gaze understanding is, most likely, not complete merely because infants can differentiate between two objects, one which has been gazed at. In daily life, we often face more nuanced, fine-grained gaze cues and must differentiate between a virtually unlimited number of potential targets.

Let us now turn to the point that often the experimenter shifts their eyes and head in synchrony [e.g., @behne2005oneyearolds]. Instead of pointing towards gaze understanding, a critic could claim that the results can be explained by face direction alone. A handful of studies approached this potential confound by separately manipulating head and eye movement. @brooks2002importance implemented a comparison between eye and head orientation and found that 14-month-olds were sensitive to open versus closed eyes. @moore1997role showed that 9-month-olds followed an agent’s gaze more when it was accompanied by a dynamic head turn in comparison to a static head turn. Investigating the 'cooperative eye hypothesis', @tomasello2007reliance implemented six conditions, in which an experimenter oriented towards the ceiling with their eyes only, head only (eyes closed), both head and eyes, or neither. They found that human infants relied more on the eye movement, while chimpanzees paid more attention to the head movement. Results of @corkum1995development, @lempers1979young, and @lempers1977development suggest that infants, at least until 19 months, struggle when eye and head direction diverge. This underlines how important the movement of the head is when interpreting the gaze of others and that the scope of infants' gaze understanding might be less comprehensive than previously assumed.

Next, let us consider the classical response measure of collecting looking times. Though previous studies using the above-mentioned methods suggest that young infants can align their visual attention to another’s line of sight, it does not necessarily include understanding the intentions or the content of the other agent’s gaze. Infants could simply attune their orientation or be attracted by others’ gaze without processing what exactly the other is seeing [cf. Butterworth & Jarrett's ecological and geometric mechanism, @butterworth1991minds]. Arguably, this is not sufficient to claim an understanding of visual perspectives, because the “target” or “object of representation” is not necessarily specified [@perner2003perspective, p. 358]. To fulfill this criterion, @moll2006level used an active behavioral choice as a response measure and tested infants' understanding of the intentions behind gaze. Infants interactively searched for the target object, which was either occluded or visually accessible. They found that 24-month-olds adaptively handed over the target object, while 18-month-olds still struggled to complete this task. In a hiding game with two search locations, @povinelli1997exploitation found that two-year-olds performed at chance level when using gaze as a cue to locate the reward, while three-year-olds passed the task. However, in a similar object choice paradigm with two containers, @behne2005oneyearolds investigated whether infants understand the communicative intent behind pointing and gaze cues. In contrast to @povinelli1997exploitation, they found that already 14-month-olds used the agent’s cues to select an object. In conditions with absent-minded ‘cues’, infants performed around chance. The specific scope of children’s gaze understanding and whether correct looking translates into meaningful active behavior deserve further investigation.

Finally, presenting two objects, focusing on group-level analysis, and passing versus failing a task drastically limit the level of detail with which we can gauge children’s gaze understanding. Undoubtedly, children can vary in the onset of their social-cognitive abilities: these skills develop over time, with experience, and on a continuum. @birch2017perspectives argued that individuals can differ in their frequency (or propensity) and accuracy of applying social cognition in their daily lives. Measuring individual differences in this domain is especially important to understand the underlying processes and to disentangle the impact of environmental influences (scaffolding) and other cognitive abilities [@birch2017perspectives].

@@

The present paper is structured in the following way. In Study 1, we explore gaze understanding across the lifespan. We assess the developmental trajectory and individual differences in this fundamental social-cognitive ability in a sample of 3- to 80-year-olds. In Study 2, we propose a computational cognitive model of gaze understanding. This model helps us to explain where individuals vary and which mechanisms participants use to locate an agent’s attentional focus. In Study 3, we revisit a fundamental assumption of our computational gaze model by experimentally isolating the vector estimation component of gaze understanding. Furthermore, we investigate the relationship between gaze understanding and other (social-) cognitive processes.

@@

We collected data from a remote child, teenager and adult sample, reaching from 3 to 80 years of age (see Supplements for further details). The child and teenager sample consisted of `r sum(lifespan_sample$n_total[lifespan_sample$age_group < 18])` participants. We recruited participants via an internal database consisting of families living in Leipzig, Germany, who volunteered to participate in child development studies and indicated an interest in online studies. Participants came from ethnically homogeneous, mixed socioeconomic backgrounds with mid to high parental education levels. They lived in an industrialized, urban Central-European context in a mid-size German city (∼600,000 inhabitants; median individual monthly net income ~ 1,600€ as of 2021). Most were raised monolingually in a nuclear two-generational family setting. Information on demographics and socioeconomic status was not formally recorded on a participant level.

@@

Our goal was to describe the inferential process of gaze understanding. We quantified how well our gaze model explained the gaze understanding process by comparing it to two alternative models that make different assumptions of which information participants use and where they consequently click to locate the target. Our modeling framework consisted of three mutually exclusive models: (1) a gaze model, (2) a random guessing model, and (3) a center bias model. We gauged which model can best explain our data by conducting model comparisons. All cognitive models were implemented in WebPPL [@goodman2014design].

@@

We found clear support for our gaze model, both in children as well as adults.  The Bayes Factors we computed via the marginal likelihood of the data strongly favored our gaze model compared to the center bias model ($logBF_{10}$ = `r gazemodel_comparison_bf$bf_inf_center`) and the random guessing model ($logBF_{10}$ = `r gazemodel_comparison_bf$bf_inf_guess`).
When correlating the observed data across all target positions with the predictions of the three models, we found a high similarity for the gaze model: *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_pred"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_pred"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_pred"]`], while the correlations with the alternative models were substantially smaller (center bias model: *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "centermodel_pred"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "centermodel_pred"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "centermodel_pred"]`]; random guessing model: *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "randommodel_pred"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "randommodel_pred"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "randommodel_pred"]`]). The developmental trajectory of the estimated gaze model parameter also correlated highly with the observed data: *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_param"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_param"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_param"]`]. The age effects in Study 2 largely replicated those of Study 1.

@@

Our findings from Study 1 showed individual differences and a developmental change in gaze understanding across the lifespan To answer what develops with age and how participants differ from one another, we presented a formal cognitive model of gaze understanding. Participants are modeled to observe all available gaze information, integrate it from both the agent's eyes, and consequently arrive at the attentional focus of the agent: the target location. We assume the basic process of gaze understanding to be the same across the lifespan, though individuals become increasingly precise with age. By conducting model comparisons, we could rule out that participants' responses can be explained by random guessing or a center bias.

@@

A limitation of our model is that we cannot disentangle how much of the participants' uncertainty comes from a noisy estimate of the agent's attentional focus and how much is due to imprecise clicking (e.g., wanting to click somewhere but experiencing motor issues at aiming, adding random noise to the click).

A critical feature of our model is that it assumes gaze understanding to rely on vector estimation. In other words, subjects are modeled to calculate pupil angles which serve as gaze vectors to point to the attentional focus point of an agent. Even though this vector estimation component is a rather physical, geometrical calculation, it still happens in a social context. In the first place, one must interpret the agent's eyes as a relevant social stimulus. Therefore, our computational model describes gaze understanding as a particular form of vector estimation in a social context.

@@

We previously presented a computational cognitive model of gaze understanding. Our model relies on the perhaps unexpected assumption that vector estimation is a crucial component of gaze understanding. In model comparisons, we found overwhelming support for this model in children and adults. Now, we wanted to test this assumption experimentally. We were interested in the degree to which vector estimation is a part of gaze understanding. Additionally, we investigated whether there is more to gaze understanding than the physical vector estimation component. To answer this question, we assessed the relationship between gaze understanding and other social-cognitive abilities. We reasoned that gaze understanding is integral to social interaction, which is most likely learned in the social world. The positive link between TANGO and family-level variables like number of siblings [@prein2023tango] underlines this. Therefore, it seems reasonable to expect correlations between gaze understanding and other social-cognitive abilities.

First, we aimed to experimentally isolate the vector estimation component of the TANGO. We designed a new non-social vector estimation task that shared all crucial design features of the TANGO. Second, we assessed children's social-cognitive abilities by administering a Theory of Mind (TOM) task battery, comprising four tasks from the ToM scale by Wellman and Liu [@wellman2004scaling] and two additional perspective-taking tasks [@flavell1981development; @flavell1981younga].

@@

Children were tested in a quiet room in their kindergarten. An experimenter guided the child through the study. Since our research questions related to individual differences and we wanted maximum control of extraneous participant variables, we employed a within-subjects study design. All participants performed the following tasks in a fixed order: (1) non-social vector estimation task, (2) ToM task battery, (3) TANGO. Several reasons motivated this decision. First, we decided on a fixed order to be able to compare participants' performance straight-forwardly with each other. Second, to increase participant engagement and decrease fatigue or fuzziness, we switched from a tablet task to tasks with personal interaction back to a tablet task. Third, we showed the non-social vector estimation task before the TANGO so that participants would not be biased to interpret the presented stimuli as "eye- /"agent-like".

@@

We administered four tasks from the @wellman2004scaling ToM scale. We excluded three tasks: the Diverse Desires task in order to avoid ceiling effects; and both tasks involving emotions (Belief Emotion and Real-Apparent Emotion), as we aimed at assessing the "cold, cognitive" (as compared to the "emotional") aspects of social cognition. Instead, we added two perspective-taking level-2 tasks [@flavell1981development; @flavell1981younga]. We added the perspective-taking tasks (1) with the aim of increasing the variablility we can capture between individuals, and (2) since we hypothesized that perspective-taking would rely on similar mechanisms than gaze understanding. The dependent variable was the aggregate score of all solved ToM tasks (see Supplements for further detail). Additionally, we investigated if gaze understanding was more strongly associated with the two perspective-taking tasks compared to the other ToM tasks, as perspective-taking seems most closely theoretically related to gaze understanding (i.e., in both cases the participant is asked to judge another person's point of view).

@@

Regarding the relationship between the two vector estimation tasks and the ToM measures, we could imagine two possible scenarios: (A) If gaze understanding recruits a general social-cognitive ability beyond vector estimation, we expected that gaze understanding and ToM measures would correlate more strongly with each other than non-social vector estimation and ToM measures. (B) If gaze understanding relies purely on task-specific processes, then the correlation between gaze understanding and ToM measures would be comparable to the correlation between non-social vector estimation and the ToM measures. For the association between the aggregate ToM scores and the gaze understanding / non-social vector estimation tasks, we used *Spearman's* rank correlation coefficients.

@@

Furthermore, to estimate which components best explain the gaze understanding score, we conducted a model comparison with GLMMs predicting the mean imprecision in gaze understanding by age, imprecision in non-social vector estimation, the ToM aggregate score, or the aggregate of the two perspective-taking tasks (subset of ToM battery; example of model notation in `R: tango_mean ~ age_centered + magnet_scaled + perspective_scaled`). We wanted to assess whether the ToM aggregate score or the singled-out perspective-taking score added additional explanatory value when predicting the gaze understanding score. The outcome variable was modeled by a lognormal distribution.

@@

Both the TANGO and the two applied perspective-taking tasks can be seen as instances of *visuospatial perspective-taking.* According to @surtees2013similarities, perspective-taking tasks involve three components: (1) a Self (i.e., the perspective-taker), (2) the Other (i.e., a target perspective that gets judged), and (3) an Object (which is taken into perspective).
One can roughly categorize them into two groups.

Level 1 perspective-taking, also termed visual accessibility judgment, is concerned with the visibility of objects from a particular viewpoint. In these tasks, participants judge whether an object is located before or behind an agent and if this agent can see it [@surtees2013similarities]. Usually, participants can solve these tasks independently from another's agent frame of reference and do not need to apply mental rotation (not matching the "mentalizing criterion"; [@quesque2020theoryofmind]).

Level 2 perspective-taking, however, does rely on a mental transformation of oneself into the space of another agent. In these tasks, participants judge how (visually or conceptually) the world looks like for another person or whether an object is to their right or left. This presumably entails a simulative transformation of one's own body schema [@erle2017grounded]. Additionally, participants must understand that different viewpoints lead to different perspectives [@flavell1981younga]: two agents can see the same object "in different, incompatible, fine-grained ways" [@rakoczy2022foundations].

In short, Level 1 perspective-taking addresses *what* an agent can see, while Level 2 perspective-taking addresses *how or where* they see it [@moll2006level]. The TANGO falls into the first category - however, in contrast to existing research, on a continuous scale. The two perspective-taking tasks we administered within the ToM scale fall into the second category.

What drives the connection between the TANGO and the two other perspective-taking tasks? Both tasks tap into the ability to judge what another person sees. In the TANGO, participants must assume that another agent perceives and looks at the target in order to locate it. The here-used perspective-taking tasks add a layer by asking how one's perspective differs from another person's and how exactly the other person sees an object. In an intriguing series of experiments, @michelon2006two assessed the different processes that participants used to master Level 1 (visibility of objects) and Level 2 (object's location relative to another agent) perspective-taking tasks. Their results suggest that participants rely on line-of-sight tracing in the former case and perspective transformation in the latter case. According to [@michelon2006two], **line-of-sight tracing** presumably consists of (1) locating the avatar, (2) locating the target, and (3) drawing a line from the agent to the target. Note how our gaze model in Study 2 shares the underlying idea of connecting points in space via a line.

Perspective transformation presumably adds complexity and consists of: "(1) locate the avatar in an egocentric space, (2) perform a perspective transformation so that one's imagined position matches the position of the avatar, (3) locate the target object in the transformed spatial representation of the scene, and (4) read off the coordinates of the object" [@michelon2006two]. Therefore, the processes to solve Level 1 perspective-taking tasks might be computationally lighter since there is no need to adapt the other person's reference frame. Still, the assumed processes overlap, which could explain the correlation between the TANGO and the administered perspective-taking tasks.

Why doesn't gaze understanding correlate with the other ToM tasks? While ToM tasks and the TANGO share the social context, the cognitive processes needed to solve each task might vary. As @rakoczy2022foundations reflected, perception-goal psychology (which includes gaze understanding) comprises understanding that others see different objects or pursue different goals. However, this ability does not necessarily entail understanding more complex meta-representational aspects; for example, understanding that mental states can be mutually incompatible, false, or involve fine-grained aspectual information.

>>>>

abstract: |
  Observing eye gaze is fundamental for many social-cognitive abilities, for example, when judging what another agent can or cannot know. While the emergence of gaze following in infancy has been thoroughly studied, we know little about the developmental trajectory throughout childhood and adolescence. In Study 1, we examine gaze understanding across the lifespan. In Study 2, we formalize the process of gaze understanding in a computational cognitive model that allows us to conceptualize individual differences in a psychologically meaningful way. The key parameter in our model is an inferential component, which describes how accurately a participant infers a target location based on an agent's gaze. In Study 3, we test a fundamental assumption of our gaze model by studying vector estimation in a non-social setting. Lastly, we explore the relationship between other social-cognitive abilities and gaze understanding. Our results suggest a steep learning curve in gaze understanding during the preschool years. Children become more precise in locating the attentional focus of an agent, while precision levels then stay comparably stable throughout adulthood. Our gaze model estimates reliable individual parameter values and recovers signature patterns in the data. Furthermore, we found that gaze understanding is associated with non-social vector estimation and perspective-taking but not with other Theory of Mind tasks. This work illustrates how the combination of reliable tasks and formal theoretical models allows us to explore the in(ter)dependence of core social-cognitive processes in greater detail.
keywords          : "social-cognitive development, theory of mind, gaze cues, individual differences, cognitive modeling, lifespan"
wordcount         : "xxx max 8000"
floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes 
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
header-includes   :
  - \usepackage{setspace}
  - \captionsetup[figure]{font={stretch=1}}
editor_options    : 
  markdown: 
    wrap: sentence
bibliography      : "../../references.bib"
---

@@

At first glance, the debate about children’s gaze following abilities seems to be settled. On further view, however, some methodological choices in the experimental paradigms as well as the scope of infants’ abilities seem to offer room for further exploration. We see the following challenges: (1) Studies focus on infancy, leaving aside further potential developments in childhood and adolescence; (2) Gaze cues are often accompanied by a head movement in the same direction, which makes it hard to isolate the signaling power of the eyes; (3) Often, infants’ looking times are taken as a benchmark for passing a task, while it is uncertain how this would translate into active behavior; (4) Presenting two objects drastically limits the complexity of the visual input, especially when compared to our daily, visually cluttered environment; (5) Behavior is grouped into passing/failing or a comparison against chance, ignoring the accuracy of children; (6) Analyses focus on group-level averages, overlooking variation in the sample. In the following paragraphs, we will turn to each of these arguments.

While the emergence of gaze following has been well established, less is known about the developmental trajectory throughout childhood and adolescence. One possibility is that our social-cognitive ability in question is fully developed once emerged in infancy. However, many cognitive abilities develop with age [e.g., working memory, @gathercole2004structure]. Similarly, visual processing appears to improve with age. Therefore, children could potentially improve in gaze following, fine-tuning the performance of the already existing skill. The development of gaze understanding is, most likely, not complete merely because infants can differentiate between two objects, of which one has been gazed at.

In traditional gaze following studies, the experimenter often shifts their eyes and head in synchrony [e.g., @behne2005oneyearolds]. Instead of pointing towards gaze understanding, a critic could claim that the results can be explained by face direction alone. A handful of studies approached this potential confound by separately manipulating head and eye movement. @brooks2002importance implemented a comparison between eye and head orientation and found that 14-month-olds were sensitive to open versus closed eyes. @moore1997role showed that 9-month-olds followed an agent’s gaze more when it was accompanied by a dynamic head turn in comparison to a static head turn. Investigating the 'cooperative eye hypothesis', @tomasello2007reliance implemented six conditions, in which an experimenter oriented towards the ceiling with their eyes only, head only (eyes closed), both head and eyes, or neither. They found that human infants relied more on the eye movement, while chimpanzees paid more attention to the head movement. Results of @corkum1995development, @lempers1979young, and @lempers1977development suggest that infants, at least until 19 months, struggle when eye and head direction diverge. This underlines how important it is to differentiate between studying head and/or eye movement. By isolating eye movement alone, we can get a clearer understanding of infants' gaze understanding. This also makes it easier to understand the processes behind gaze understanding.

Next, let us consider the classical response measure of collecting looking times. Though previous studies using the above-mentioned methods suggest that young infants can align their visual attention to another’s line of sight, it does not necessarily include understanding the intentions or the content of the other agent’s gaze. Infants could simply attune their orientation or be attracted by others’ gaze without processing what exactly the other is seeing [cf. Butterworth & Jarrett's ecological and geometric mechanism, @butterworth1991minds]. Arguably, this is not sufficient to claim an understanding of visual perspectives, because the “target” or “object of representation” is not necessarily specified [@perner2003perspective, p. 358]. To fulfill this criterion, @moll2006level used an active behavioral choice as a response measure and tested infants' understanding of the intentions behind gaze. Infants interactively searched for the target object, which was either occluded or visually accessible. They found that 24-month-olds adaptively handed over the target object, while 18-month-olds still struggled to complete this task. In a hiding game with two search locations, @povinelli1997exploitation found that two-year-olds performed at chance level when using gaze as a cue to locate the reward, while three-year-olds passed the task. However, in a similar object choice paradigm with two containers, @behne2005oneyearolds investigated whether infants understand the communicative intent behind pointing and gaze cues. In contrast to @povinelli1997exploitation, they found that already 14-month-olds used the agent’s cues to select an object. In conditions with absent-minded ‘cues’, infants performed around chance. The specific scope of children’s gaze understanding and whether looking toward the cued location translates into meaningful active behavior deserve further investigation.

Finally, presenting two objects, focusing on group-level analysis, and passing versus failing a task drastically limit the level of detail with which we can gauge children’s gaze understanding. In daily life, we must differentiate between a virtually unlimited number of potential targets and often face more nuanced, fine-grained gaze cues. Undoubtedly, children can vary in the onset of their social-cognitive abilities: these skills develop over time, with experience, and on a continuum. @birch2017perspectives argued that individuals can differ in their frequency (or propensity) and accuracy of applying social cognition in their daily lives. Measuring individual differences in this domain is especially important to understand the underlying processes and to disentangle the impact of environmental influences (scaffolding) and other cognitive abilities [@birch2017perspectives].

@@

The present paper is structured in the following way. In Study 1, we explore gaze understanding across the lifespan. We assess the developmental trajectory and individual differences in this fundamental social-cognitive ability in a sample of 3- to 80-year-olds. In Study 2, we propose a computational cognitive model of gaze understanding. This model helps us to explain where individuals vary and which mechanisms participants use to locate an agent’s attentional focus. In Study 3, we revisit a fundamental assumption of our computational gaze model by experimentally isolating the vector estimation component of gaze understanding. Furthermore, we investigate the relationship between gaze understanding and other social-cognitive processes.

@@

We collected data from a remote child, teenager and adult sample, reaching from 3 to 80 years of age (see Supplements for further details). The child and teenager sample consisted of `r sum(lifespan_sample$n_total[lifespan_sample$age_group < 18])` participants. We recruited participants via an internal database consisting of families living in Leipzig, Germany, who volunteered to participate in child development studies and indicated an interest in online studies. Participants came from ethnically homogeneous, mixed socioeconomic backgrounds with mid to high parental education levels. They lived in an industrialized, urban Central-European context in a mid-size German city (∼600,000 inhabitants; median individual monthly net income approx. 1,600€ as of 2021). Most were raised monolingually in a nuclear two-generational family setting. Information on demographics and socioeconomic status was not formally recorded on a participant level.

@@

Our goal was to describe the inferential process of gaze understanding. We quantified how well our gaze model explained the gaze understanding process by comparing it to two alternative models that make different assumptions of which information participants use and where they consequently click to locate the target. Our modeling framework consisted of three models: (1) a gaze model, (2) a random guessing model, and (3) a center bias model. We gauged which model can best explain our data by conducting model comparisons. All cognitive models were implemented in WebPPL [@goodman2014design].

@@

We found clear support for our gaze model, both in children as well as adults.  The Bayes Factors we computed via the marginal likelihood of the data strongly favored our gaze model compared to the center bias model ($logBF_{10}$ = `r gazemodel_comparison_bf$bf_inf_center`) and the random guessing model ($logBF_{10}$ = `r gazemodel_comparison_bf$bf_inf_guess`).
When correlating the observed data across all target positions with the predictions of the three models, we found a high similarity for the gaze model: *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_pred"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_pred"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_pred"]`], while the correlations with the alternative models were substantially smaller (center bias model: *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "centermodel_pred"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "centermodel_pred"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "centermodel_pred"]`]; random guessing model: *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "randommodel_pred"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "randommodel_pred"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "randommodel_pred"]`]). As can be seen in Figure \@ref(fig:fig2)E, the gaze model could recover the U-shaped pattern that we observed our our data. The developmental trajectory of the estimated gaze model parameter also correlated highly with the observed data: *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_param"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_param"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_param"]`]. The age effects in Study 2 largely replicated those of Study 1.

@@

Our findings from Study 1 showed individual differences and a developmental change in gaze understanding across the lifespan. To answer what develops with age and how participants differ from one another, we presented a formal cognitive model of gaze understanding. Participants are modeled to observe all available gaze information, integrate it from both the agent's eyes, and consequently arrive at the attentional focus of the agent: the target location. We assume the basic process of gaze understanding to be the same across the lifespan, though individuals become increasingly precise with age. By conducting model comparisons, we could rule out that participants' responses can be explained by random guessing or a center bias.

@@

A limitation of our model is that we cannot disentangle how much of the participants' uncertainty comes from a noisy estimate of the agent's attentional focus and how much is due to imprecise clicking (e.g., wanting to click somewhere but experiencing motor issues at aiming, adding random noise to the click). However, the children participating in our study seemed rather determined in where they wanted to click and to not have issues with aiming or motor control. Therefore, imprecise clicking should not be a great concern.

A critical feature of our model is that it assumes gaze understanding to rely on vector estimation. In other words, subjects are modeled to calculate pupil angles which serve as gaze vectors to point to the attentional focus point of an agent. Even though this vector estimation component is a rather physical, geometrical calculation, it still happens in a social context. For example, one must interpret the agent's eyes as a relevant social stimulus. Therefore, our computational model describes gaze understanding as a particular form of vector estimation in a social context.

@@

We previously presented a computational cognitive model of gaze understanding. Our model relies on the perhaps unexpected assumption that vector estimation is a crucial component of gaze understanding. In model comparisons, we found overwhelming support for this model in children and adults. Now, we wanted to further explore this assumption experimentally. We were interested in the degree to which vector estimation is a part of gaze understanding. Additionally, we investigated whether there is more to gaze understanding than the physical vector estimation component. To answer this question, we assessed the relationship between gaze understanding and other social-cognitive abilities. We reasoned that gaze understanding is integral to social interaction, which is most likely learned in the social world. The positive link between TANGO and family-level variables like number of siblings [@prein2023tango] underlines this. Therefore, it seems reasonable to expect correlations between gaze understanding and other social-cognitive abilities.

First, we aimed to experimentally isolate the vector estimation component of the TANGO. We designed a new non-social vector estimation task that shared all crucial design features of the TANGO. Second, we assessed children's social-cognitive abilities by administering a Theory of Mind (TOM) task battery, comprising four tasks from the ToM scale by Wellman and Liu [@wellman2004scaling], and two additional perspective-taking tasks [@flavell1981development; @flavell1981younga].

@@

Children were tested in a quiet room in their kindergarten. An experimenter guided the child through the study. Since our research questions related to individual differences and we wanted maximum control of extraneous participant variables, we employed a within-subjects study design. All participants performed the following tasks in a fixed order: (1) non-social vector estimation task, (2) ToM task battery, (3) TANGO. Several reasons motivated this decision. First, we decided on a fixed order to be able to compare participants' performance straight-forwardly with each other. Second, to increase participant engagement and decrease fatigue or fuzziness, we switched from a tablet task to tasks with personal interaction back to a tablet task. Third, we showed the non-social vector estimation task before the TANGO so that participants would not be biased to interpret the presented stimuli as "eye- /agent-like".

@@

We administered four tasks from the @wellman2004scaling ToM scale. We excluded three tasks: the Diverse Desires task in order to avoid ceiling effects; and both tasks involving emotions (Belief Emotion and Real-Apparent Emotion), as we aimed at assessing the "cold, cognitive" (as compared to the "emotional") aspects of social cognition. Instead, we added two perspective-taking level-2 tasks [@flavell1981development; @flavell1981younga]. We added the perspective-taking tasks (1) with the aim of increasing the variablility we can capture between individuals, and (2) since we hypothesized that perspective-taking would rely on similar mechanisms than gaze understanding.

@@

For the ToM battery, the dependent variable was the aggregate score of all solved tasks (see Supplements for further detail). Regarding the relationship between the two vector estimation tasks and the ToM measures, we could imagine two possible scenarios: (A) If gaze understanding recruits a general social-cognitive ability beyond vector estimation, we expected that gaze understanding and ToM measures would correlate more strongly with each other than non-social vector estimation and ToM measures. (B) If gaze understanding relies purely on task-specific processes, then the correlation between gaze understanding and ToM measures would be comparable to the correlation between non-social vector estimation and the ToM measures. For the association between the aggregate ToM scores and the gaze understanding / non-social vector estimation tasks, we used *Spearman's* rank correlation coefficients.

@@

Furthermore, to estimate which components best explain the gaze understanding score, we conducted a model comparison with GLMMs predicting the mean imprecision in gaze understanding by age, imprecision in non-social vector estimation, the ToM aggregate score, or the aggregate of the two perspective-taking tasks (subset of ToM battery; example of model notation in `R: tango_mean ~ age_centered + magnet_scaled + perspective_scaled`). The outcome variable was modeled by a lognormal distribution. We wanted to assess whether the ToM aggregate score or the singled-out perspective-taking score added additional explanatory value when predicting the gaze understanding score. As pre-registered, we hypothesized that perspective-taking seemed most closely theoretically related to gaze understanding as in both cases the participant is asked to judge another person's point of view.

@@

Both the TANGO and the two applied perspective-taking tasks can be seen as instances of *visuospatial perspective-taking.* According to @surtees2013similarities, perspective-taking tasks involve three components: (1) a Self (i.e., the perspective-taker), (2) the Other (i.e., a target perspective that gets judged), and (3) an Object (which is taken into perspective). Often, researchers distinguish between Level 1 and Level 2 perspective-taking tasks. Level 1 perspective-taking, also termed visual accessibility judgment, is concerned with the visibility of objects from a particular viewpoint. In these tasks, participants judge whether an object is located before or behind an agent and if this agent can see it [@surtees2013similarities]. Usually, participants can solve these tasks independently from another's agent frame of reference and do not need to apply mental rotation (not matching the "mentalizing criterion"; [@quesque2020theoryofmind]). Level 2 perspective-taking, however, does rely on a mental transformation of oneself into the space of another agent. In these tasks, participants judge how (visually or conceptually) the world looks like for another person or whether an object is to their right or left. This presumably entails a simulative transformation of one's own body schema [@erle2017grounded]. Additionally, participants must understand that different viewpoints lead to different perspectives [@flavell1981younga]: two agents can see the same object "in different, incompatible, fine-grained ways" [@rakoczy2022foundations]. In short, Level 1 perspective-taking addresses *what* an agent can see, while Level 2 perspective-taking addresses *how or where* they see it [@moll2006level]. The TANGO falls into the first category - however, in contrast to existing research, on a continuous scale. The two perspective-taking tasks we administered within the ToM scale fall into the second category.

What could drive the connection between the TANGO and the two other perspective-taking tasks? Both tasks tap into the ability to judge what another person sees. In the TANGO, participants must assume that another agent perceives and looks at the target in order to locate it. The here-applied Level 2 perspective-taking tasks add a layer by asking how one's perspective differs from another person's and how exactly the other person sees an object. In an intriguing series of experiments, @michelon2006two assessed the different processes that participants used to master Level 1 (visibility of objects) and Level 2 (object's location relative to another agent) perspective-taking tasks. Their results suggest that participants rely on line-of-sight tracing in the former case and perspective transformation in the latter case. According to [@michelon2006two], line-of-sight tracing presumably consists of (1) locating the avatar, (2) locating the target, and (3) drawing a line from the agent to the target. Note how our gaze model in Study 2 shares the underlying idea of connecting points in space via a line. Perspective transformation presumably adds complexity and consists of: "(1) locate the avatar in an egocentric space, (2) perform a perspective transformation so that one's imagined position matches the position of the avatar, (3) locate the target object in the transformed spatial representation of the scene, and (4) read off the coordinates of the object" [@michelon2006two]. Therefore, the processes to solve Level 1 perspective-taking tasks might be computationally lighter since there is no need to adapt the other person's reference frame. Still, the assumed processes overlap, which could explain the correlation between the TANGO and the administered perspective-taking tasks.

@surtees2013similarities further differentiate between visual and spatial perspective-taking. While the former helps to judge if and how an agent sees an object, the latter involves judging the relative spatial locations of an agent and the object. Spatial perspective-taking can work without mental states and can, therefore, be applied to non-agentive objects with a front [@surtees2013similarities]. Interestingly, computing a line of sight does not demand the presence of another agent [@michelon2006two]. This could explain how our participants solved the non-social vector estimation task.
